---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bis620.2022

<!-- badges: start -->
[![R-CMD-check](https://github.com/wanjingc7/bis620.2022/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/wanjingc7/bis620.2022/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/wanjingc7/bis620.2022/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/wanjingc7/bis620.2022/actions/workflows/test-coverage.yaml)
[![Codecov test coverage](https://codecov.io/gh/wanjingc7/bis620.2022/branch/main/graph/badge.svg)](https://app.codecov.io/gh/wanjingc7/bis620.2022?branch=main)
[![lint](https://github.com/wanjingc7/bis620.2022/actions/workflows/lint.yaml/badge.svg)](https://github.com/wanjingc7/bis620.2022/actions/workflows/lint.yaml)
<!-- badges: end -->



The goal of bis620.2022 is to 

For the Accelerometry data: 

1. Capture spectral signature of Accelerometry data with three dimensions using Fast Fourier Transform (FFT) and also calculate frequency with time.

2. Create a visualization for movements in three dimensions with time or frequency.


For the clinical trials data of Panitumumab on head-neck cancer patients:

1. Create histograms of distribution of background information of the 
participants, including age, sex, race, diagnosis (with or without stages).

3. Create histograms of distribution of the treatment outcome: adverse events
and best treatment response.

4. Conduct logistic regressions to test whether a certain factor is prognostic
or predictive. 


## Links to coverage and Lint Results

[Links to Coverage](https://github.com/wanjingc7/bis620.2022/actions/workflows/test-coverage.yaml)

[Links to Lint Results](https://github.com/wanjingc7/bis620.2022/actions/workflows/lint.yaml)

## Installation

You can install the development version of bis620.2022 from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("wanjingc7/bis620.2022")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example Accelerometry data}
library(bis620.2022)
## basic example code

#to get FFT of dimensions and frequency from time
data("ukb_accel")
ukb_accel |> spectral_signature()

#to create visualization of movement with time
ukb_accel |> accel_plot()

#to create visualization of logarithm transformed FFT
#movement with frequency
ukb_accel |> spectral_signature(take_log = T) |> accel_plot()
```


```{r example Clinical Trials data}
#only a few examples shown here, to see full list of examples, please
#check the vignette.

data("full_data")

#plot age distribution of participants
plot_age(df = full_data, subject_id_col = "SUBJID", age_col = "AGE",
         treat_col = "ATRT", age_interval = 5)


#plot adverse events distribution of participants
get_ae(df = full_data, ae_col = "AETERM", treat_col = "ATRT", top = 30)

#conduct a logstic regression to see if age is prognostic
prognostic_var(df = full_data, subject_id_col = "SUBJID", treat_col = "ATRT",
               control_val = "Chemotherapy", death_col = "DEATH_FLAG", 
               var_col = "AGE")

#conduct a logstic regression to see if age is predictive
predictive_var(df = full_data, subject_id_col = "SUBJID",
               death_col = "DEATH_FLAG", treat_col = "ATRT", var_col = "AGE",
               interaction = T)
```
